Finally, a functional class scoring method for enrichment analysis of Infinium DNA methylation data

### 2023-11-17 figs/GSE158422_sensitivity.Rmd
sensitivity done

@@

sizes=c(1000,5000,10000,20000,50000,100000,200000,300000,350000,400000,450000,500000)
nsigs <- lapply(sizes, function(z) {
  dmsig <- head(dm,z)
  dmsigup <- rownames(subset(dmsig,logFC>0))
  dmsigdn <- rownames(subset(dmsig,logFC<0))
  gsaup <- gsameth(sig.cpg=dmsigup, all.cpg=rownames(dm), collection=gs_entrez)
  rownames(gsaup) <- paste(rownames(gsaup),"UP")
  gsadn <- gsameth(sig.cpg=dmsigdn, all.cpg=rownames(dm), collection=gs_entrez)
  rownames(gsadn) <- paste(rownames(gsadn),"DN")
  nsig <- nrow(subset(rbind(gsaup,gsadn),FDR<0.05))
  return(nsig)
})
nsigs <- unlist(nsigs)
names(nsigs) <- sizes
par( mar = c(5.1, 6.1, 4.1, 2.1) )
barplot(nsigs,horiz=TRUE,las=1,xlab="no. significant sets")
mtext("no. probes selected", 2, 4)

sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))

length(sigup)
length(sigdn)

gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
rownames(gsadn) <- paste(rownames(gsadn),"DN")
ngsig <- nrow(subset(rbind(gsaup,gsadn),FDR<0.05))

tic()
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
rownames(gsadn) <- paste(rownames(gsadn),"DN")
toc() # 12.9, 12.6, 11.5

>>>>

dim(dm)
nrow(subset(dm,adj.P.Val<0.05 & logFC>0))
nrow(subset(dm,adj.P.Val<0.05 & logFC<0))

length(unique(unlist(strsplit(subset(dm,adj.P.Val<0.05 & logFC>0)$UCSC_RefGene_Name,";"))))
length(unique(unlist(strsplit(subset(dm,adj.P.Val<0.05 & logFC<0)$UCSC_RefGene_Name,";"))))

length(unique(unlist(strsplit(dm$UCSC_RefGene_Name,";"))))

@@

sizes=c(1000,5000,10000,20000,50000,100000,200000,300000,350000,400000,450000,500000)
nsigs <- lapply(sizes, function(z) {
  dmsig <- head(dm,z)
  dmsigup <- rownames(subset(dmsig,logFC>0))
  dmsigdn <- rownames(subset(dmsig,logFC<0))
  gsaup <- gsameth(sig.cpg=dmsigup, all.cpg=rownames(dm), collection=gs_entrez)
  rownames(gsaup) <- paste(rownames(gsaup),"UP")
  gsadn <- gsameth(sig.cpg=dmsigdn, all.cpg=rownames(dm), collection=gs_entrez)
  rownames(gsadn) <- paste(rownames(gsadn),"DN")
  nsig <- rownames(subset(rbind(gsaup,gsadn),FDR<0.05))
  return(nsig)
})

names(nsigs) <- sizes
nsigs

par( mar = c(5.1, 6.1, 4.1, 2.1) )
barplot(unlist(lapply(nsigs,length)),horiz=TRUE,las=1,xlab="no. significant sets") ; grid()
mtext("no. probes selected", 2, 4)

#
FCLIM <- c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)
ngsig <- lapply(FCLIM,function(F) {
  sigup <- rownames(subset(dm,logFC > F & adj.P.Val < 0.05))
  sigdn <- rownames(subset(dm,logFC < -F & adj.P.Val < 0.05))
  gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
  rownames(gsaup) <- paste(rownames(gsaup),"UP")
  gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
  rownames(gsadn) <- paste(rownames(gsadn),"DN")
  ngsig <- rownames(subset(rbind(gsaup,gsadn),FDR<0.05))
  return(ngsig)
})

names(ngsig) <- FCLIM
ngsig

par( mar = c(5.1, 6.1, 4.1, 2.1) )
barplot(unlist(lapply(ngsig,length)),horiz=TRUE,las=1,xlab="no. significant sets",xlim=c(0,60)) ; grid()
mtext("fold change threshold", 2, 4)

sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
tic()
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
rownames(gsadn) <- paste(rownames(gsadn),"DN")
toc() # 12.9, 12.6, 11.5

### 2023-11-17 manuscript/manuscript.Rmd
sensitivity done

![Figure 1. Precision, recall and F1 for enrichment tests with simulated data.](../figs/fig1.png)

@@

![Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests over a range of group sizes and delta values. One hundred genes per set.](../figs/fig2.png)

@@

Next, GSA and LAM methods were applied to compare paired normal and cancer methylation data.

* Chart results with 37 pax.

* Show sensitivity analysis with range of samples from 3 to 30.

* LAM has better performance than GSAmeth

* Could be due to (1) filtering genes without annotations, and (2) FDR not applied for sets without members.

* Can make changes to the GSAmeth codes to fix the problems.

### Comparision of AL method to existing pathway enrichment approaches

>>>>

![Figure 1. Precision, recall and F1 for enrichment tests with simulated data. Gene set sizes vary between 20 and 100 genes per set.](../figs/fig1.png)

@@

![Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests with simulated data over a range of group sizes and delta values. One hundred genes per set.](../figs/fig2.png)

@@

Next, GSA and LAM methods were applied to compare paired normal and cancer methylation data from 37 lung squamous cell carcinoma patients.
From the 839,473 probes examined, 473,572 probes were considered significant at 5% FDR.
Of these, 119,785 probes were significantly hypermethylated and 353,787 were significantly hypomethylated.
These significantly hypermethylated probes mapped to 16,520 unique genes, while the hypomethylated probes mapped to 24,114 genes.
These numbers are large, considering that there are 26,219 genes represented on this array.

LAM resulted in 358 significant Reactome pathways at FDR<0.05 while there were just 21 with GSA (*Figure 3A*).
The overlap of significant pathways was unexpectedly small, with just five common pathways.

Given GSA found relatively few results

Considering the number of significant probes was very large we were concerned that

We were curious about whether LAM method had a bias towards larger gene sets as the recall of this
method is higher with larger gene sets.
Unexpectedly, pathways identified by LAM were collectively smaller than GSA (*Figure 3B*).
Although algorithmic accuracy cannot be inferred from the types of pathways identified in real data,
top ranked results from LAM appeared to be more related to cell differentiation, identity and development (*Figure 3C*),
as compared to GSA (*Figure 3D*).

![Figure 3. Pathway enrichment analysis comparing normal and cancer samples using LAM and GSA methods. (A) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from LAM and GSA results. (B) Sizes of significant gene sets found with LAM and GSA methods. (C) Significant LAM pathways with largest S distance, a proxy for effect size. (D) Top ranked significant GSA pathways ranked by fold enrichment.](../figs/fig3.png)

To compare the sensitivity of these methods, sample sizes were randomly downsampled, analysed
as above and the significant pathways were compared to the results from the full
group of 37 patients.
This process was repeated 50 times and the results shown in (*Figure 4*).
Pathways that were identified in the smaller and full group were termed "consistent",
while pathays that were identified in the smaller but not in the full group were
termed "inconsistent".
LAM was able to detect 300 out of 358 consistent pathways with a sample size of 10,
while GSA detected just 4 of 21 at this sample size.
LAM identified more inconsistent pathways than GSA, but the proportion of
inconsistent findings was lower across all sample sizes.
At a sample size of 10, the proportion of inconsistent pathways was 10% for LAM and
60% for GSA.

![Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.](../figs/fig4.png)

This result suggests that LAM has superior sensitivity to detect differentially methylated
pathways in real cancer data.

## LAM and GSA integrative analysis with RNA expression

* Separate promoter and gene body integration

## Exploration of methylation pathways altered by ART

## Exploration of methylation pathways altered by elevated homocysteine

##
